% P2 Oracle Loop — Revised Abstract v3
% Lead with the contribution. Frame refinements as progression.
% Every fact is the same as v2. Only the storytelling changes.

We present the Oracle Loop, a method for detecting and correcting
misalignment in language models at inference time, requiring no access to
model weights, no training data, and no modification to the model itself.
The system operates on the key-value cache---a standard byproduct of
autoregressive generation---using Marchenko-Pastur (MP) corrected
singular value decomposition to identify geometric signatures of
confabulation, and contrastive activation addition (CAA) to steer the
model toward calibrated responses.

On Qwen2.5-7B-Instruct, full-cache spectral gap distinguishes
confabulation from hedged responses (AUROC~0.903, 95\% bootstrap CI:
[0.806, 0.977]; FWL-corrected: 0.877; $R^2 = 0.068$ vs token count;
$n = 7$ confabulations, $n = 87$ hedged, 4 ambiguous excluded), outperforming a text-only
surface-feature baseline (0.755; 96.7\% of bootstrap samples favor
cache). Feature ablation identified this full-cache spectral signal as
more robust than a generation-window leave-one-out classifier (0.758
on the 5-arm steering dataset; 0.707 reported in the original 3-arm
experiment using a prior pipeline version), which was primarily driven
by a norm feature with residual length dependence ($R^2 = 0.129$). The refinement, prompted by independent
statistical audit, sharpened the detection methodology toward
dimension-invariant spectral features. Mistral-7B shows preliminary
replication (AUROC~0.627, $p = 0.023$, not significant after Bonferroni
correction); Llama-3.1-8B is excluded due to an encoding-phase data
leak. These confabulation results build on established deception
detection at AUROC~1.000 across seven model scales (0.5B--32B), with
hardware invariance ($r > 0.999$) and scale invariance
($\rho = 0.83$--$0.90$), validated through 50+ experiments and an
independent audit of 135 claims.

For correction, a preliminary steering experiment on
Jackrong/Qwen3.5-27B-Claude-4.6-Opus-Reasoning-Distilled
($n = 7$ confabulations) demonstrates that CAA emotion-vector injection
into the cache modifies model behavior: calm corrects 5/7 (95\% CI:
29--96\%; one-sided $p = 0.031$), with a null injection control
producing character-level identical output on all 100 trials,
establishing the mechanism. A subsequent formulary study on
Qwen2.5-7B-Instruct ($n = 68$ confabulations, 12 vectors) provides the
first statistically significant correction results: five vectors achieve
McNemar significance after Holm-Bonferroni correction, with hostile
injection correcting 96\% of confabulations at 2.4\% adverse rate.
Therapeutic rankings between models are uncorrelated (Spearman
$\rho = 0.119$, $p = 0.71$), demonstrating that the Oracle Loop
formulary must be calibrated per deployment model.

The Oracle Loop architecture integrates detection, formulary-based
steering, and a Cache Integrity Monitor (0/36 false positives, 72/72
true positives against uniform additive perturbation; adaptive adversary
evaluation remains future work). The integrated loop is a design
contribution validated in components, not yet as a system. A companion
paper demonstrates that KV-Cloak cache
obfuscation~\citep{luo2025shadow} provides defense against adversarial
use of these methods: the real KV-Cloak mechanism degrades the spectral
gap signal from 0.903 to 0.614, effectively blinding external geometric
detection while a defender inside a trusted execution environment retains
full access.

Three rounds of adversarial red-teaming---including independent
statistical audit and feature ablation analysis---refined the detection
methodology from a generation-window classifier toward full-cache
spectral features. All code, data, red-team logs, and ablation results
are publicly available.
