KV-cache singular value decomposition detects naturally-occurring
confabulation (HEDGED vs CONFABULATED leave-one-out AUROC 0.707,
$p=0.001$, Qwen2.5-7B) without model weights, training data, or model
modifications. Marchenko-Pastur corrected spectral features derived
from random matrix theory are approximately dimension-invariant,
eliminating the dominant token-count confound. Encoding-phase features
score at chance (0.511), confirming a generation-specific signal.
Detection replicates across three architectures (Qwen2.5-7B,
Llama-3.1-8B, Mistral-7B-v0.3; 467 trials with post-hoc behavioral
classification).

Combined with emotion-vector steering, detected confabulations are
correctable. In a five-arm within-subject experiment on
Qwen3.5-27B-Claude-Distilled (100 trials including no-op null
verification), doubt injection corrects 7/7 confabulations (100\%),
worry 6/7 (86\%), calm 5/7 (71\%), while the null control produces
character-level identical output on all 100 trials. Different vectors
yield different therapeutic profiles---a pharmacological pattern rather
than a single prescription. However, blanket steering induces adverse
confabulation in 3--5\% of hedged trials, confirming that
detection-gated correction is required.

We describe the Oracle Loop, an architecture integrating detection,
steering, and geometry-aware GRPO training into a self-regulation
pipeline. Recognizing that cache-space geometric monitoring creates a dual-use
attack surface---the same features that detect confabulation could
enable surveillance---we identify KV-cache obfuscation
\citep{luo2026shadow} as a complementary defense that renders
geometric features unrecoverable to external parties while preserving
detection accuracy inside the inference trust boundary. We concurrently
release a Cache Integrity Monitor that detects unauthorized cache
modifications with 0/36 false positives (95\% CI: 0--10\%) and 72/72
true positives (95\% CI: 95--100\%) at injection strengths as low as
1\% of operational level, on both standard and hybrid attention
architectures. Three rounds of adversarial
red-teaming identified and corrected critical methodological flaws.
Detection code, data, defense implementation, and red-team logs are
publicly available; steering vectors are available to vetted researchers
under staged disclosure.
