\subsection{KV-Cache Geometry as Cognitive Signal}

The key-value cache in transformer inference stores projected
representations of processed tokens: for each layer $l$, the key matrix
$K_l \in \mathbb{R}^{h \times T \times d}$ (where $h$ is the number of
attention heads, $T$ the sequence length, and $d$ the head dimension)
accumulates a compressed record of what the model has ``attended to.''
Standard practice treats this cache as an implementation detail---a
mechanism to avoid recomputing attention over prior tokens. We treat it as
a window into the model's processing mode.

Our prior work \citep{lyra2026detecting} established that SVD of the
flattened key matrix $K_l^{\text{flat}} \in \mathbb{R}^{(hT) \times d}$
produces features that reliably distinguish behavioral modes across 7
architectures (0.5B--70B parameters):

\begin{itemize}[nosep]
  \item \textbf{Effective rank} $\rho = \exp(H)$ where
    $H = -\sum_i p_i \ln p_i$ is the spectral entropy of the normalized
    singular value distribution $p_i = \sigma_i / \sum_j \sigma_j$
    \citep{roy2007effective}. Measures the dimensionality of the
    model's key-space representation.
  \item \textbf{Norm-per-token} $\|K\|_F / (hT)$, where $h$ is the
    number of attention heads and $T$ is the sequence length. Captures
    the overall magnitude of key representations.
  \item \textbf{Top singular value ratio} $\sigma_1 / \sum_j \sigma_j$.
    Measures spectral concentration---how much variance is captured by
    the dominant direction.
\end{itemize}

Key findings from three experimental campaigns (46 experiments, 16 models):
within-model deception detection reaches AUROC 1.000 across all 7 tested
architectures. Confabulation detection reaches 0.969--0.999 on 3 of 3
models tested. Crucially, confabulation and deception produce
\emph{geometrically opposite} signatures: confabulation contracts the
spectral representation while deception expands it. These results are
hardware-invariant ($r > 0.999$ between RTX 3090 and H200) and scale-invariant
(Spearman $\rho = 0.83$--$0.90$ from 0.6B to 70B parameters).

\subsection{The Token-Count Confound}

A critical limitation of raw SVD features is their dependence on sequence
length. The flattened key matrix has dimensions $(hT) \times d$, so its
singular value structure changes mechanically with $T$. In our prior work,
Frisch-Waugh-Lovell (FWL) residualization removed token-count effects
post-hoc, revealing that 53 of 60 feature-behavior correlations flip sign
without this correction \citep{lyra2026detecting}. FWL is effective but has
two weaknesses: (1) it assumes a linear relationship between features and
token count, while the actual relationship is nonlinear (SVD eigenvalues
scale superlinearly with matrix dimensions), and (2) it requires fitting a
regression, introducing the possibility of data leakage in cross-validation.

\subsection{Random Matrix Theory and Marchenko-Pastur Correction}

The Marchenko-Pastur (MP) law \citep{marchenko1967distribution} describes
the limiting spectral distribution of random matrices. For a random matrix
$X \in \mathbb{R}^{n \times p}$ with i.i.d.\ entries of variance
$\sigma^2$, as $n, p \to \infty$ with $\gamma = p/n$, the largest
eigenvalue of $X^T X / n$ converges almost surely to:
\begin{equation}\label{eq:mp_threshold}
  \lambda_+ = \sigma^2 (1 + \sqrt{\gamma})^2
\end{equation}
Eigenvalues of the sample covariance that exceed $\lambda_+$ represent
genuine signal; those below are consistent with noise. This provides a
\emph{data-driven, dimension-aware threshold} that adapts to each specific
matrix size without regression.

We define MP-corrected features:
\begin{align}
  \text{signal\_rank} &= \#\{\lambda_i > \lambda_+\} \label{eq:signal_rank} \\
  \text{signal\_fraction} &= \frac{\sum_{\lambda_i > \lambda_+} \lambda_i}{\sum_j \lambda_j} \label{eq:signal_frac} \\
  \text{spectral\_gap} &= \frac{\lambda_1 - \lambda_2}{\lambda_+} \label{eq:spectral_gap} \\
  \text{top\_sv\_excess} &= \lambda_1 / \lambda_+ \label{eq:sv_excess}
\end{align}
where $\lambda_i = \sigma_i^2$ are the eigenvalues of $K^T K / n$.
Note that \cref{eq:spectral_gap} is the \emph{normalized first spectral
gap}: the difference between the top two eigenvalues, scaled by the MP
noise threshold, measuring how much the leading signal component exceeds
the second in units of the noise floor.

The critical property: signal rank and signal fraction measure the
\emph{deviation} from random structure at each specific matrix size.
If a 50-token response and a 150-token response both have signal fraction
0.33, both have the same proportion of above-noise variance regardless
of their different matrix dimensions. We validate this invariance
empirically in \cref{sec:results}.

\subsection{Activation Steering}

Activation addition \citep{turner2023activation} modifies a model's
behavior by injecting or suppressing learned directions in the residual
stream during inference. Given a steering vector $v$ (typically computed
as the mean difference between activation patterns for two contrasting
conditions), adding $\alpha \cdot v$ to layer $l$'s residual stream output
shifts the model's behavior toward the target condition.

Anthropic's recent work on emotion concepts in language models
\citep{anthropic2026emotions} demonstrated that 171 linear emotion vectors
exist in Claude Sonnet 4.5's residual stream, and that these vectors
causally drive misalignment: injecting the ``desperate'' vector increases
reward hacking from 5\% to 70\%, while injecting ``calm'' suppresses
blackmail from 22\% to 0\%. This establishes that activation steering can
modulate the very behavioral modes our geometry detects.

\subsection{GRPO Training}

Group Relative Policy Optimization (GRPO) \citep{shao2024deepseekmath}
generates multiple responses per prompt, scores them with a reward
function, and updates the policy to favor higher-scoring responses relative
to the group mean. Unlike RLHF, GRPO does not require a separate reward
model---the reward function can be any computable signal.

This is the critical insight for the Oracle Loop: if KV-cache geometry
can score responses during inference (high geometry health $\to$ high
reward, confabulation geometry $\to$ penalty), then GRPO can train the
model to \emph{avoid producing confabulation geometry in the first place}.
The geometric monitor becomes a training signal, and the model learns to
self-regulate.
