We have presented a detection-and-intervention architecture for runtime
confabulation monitoring in language models, using KV-cache geometry as
both detection signal and steering input.

The detection experiment shows that Marchenko-Pastur corrected spectral
features distinguish naturally-occurring hedged and confabulated responses
with LOO AUROC of 0.707 ($p = 0.001$) on Qwen2.5-7B, using
generation-phase features only, with encoding features at chance. The MP
correction eliminates the dominant token-count confound through random
matrix theory rather than regression.

The steering experiment shows that detection-gated injection corrects
confabulation on Qwen3.5-27B-Distilled, with graded efficacy across
three emotion vectors: doubt (7/7, 100\%), worry (6/7, 86\%), calm
(5/7, 71\%), while a null injection control produces identical output
on all 100 trials. Cross-model validation on Qwen2.5-7B-Instruct
(350 trials, 68 baseline confabulations) confirms hostile as a
near-universal corrector (95.6\%, McNemar $p < 0.001$) and reveals
therapeutic inversion: vectors therapeutic on the distilled model become
iatrogenic on the base model. But blanket steering induces adverse
confabulation in 3--5\% of hedged trials, confirming that detection is
not optional---it is the prerequisite for safe correction.

Three rounds of adversarial red-teaming identified and corrected critical
methodological flaws, including a row-permutation invariance bug that would
have silently invalidated all MP features. We report these not as
embarrassments but as demonstrations that adversarial methodology is
essential for claims in this space.

The Oracle Loop creates dual-use risks at two levels. Cache injection
could suppress safety-relevant hedging; our Cache Integrity Monitor
(\cref{sec:cache_defense}) detects all tested injection attacks with
0/36 false positives and 72/72 true positives, including at 1\% of
operational strength. More broadly, the same geometric features that
detect confabulation could enable surveillance or persona
identification if applied to intercepted caches. KV-cache obfuscation
\citep{luo2026shadow} addresses this by rendering geometric features
unrecoverable to external parties while preserving detection accuracy
inside the trust boundary (\cref{sec:kv_cloak}). We advocate that any
deployment of cache-geometric monitoring include obfuscation as a
mandatory companion: the pharmacy must be locked.

A natural extension is to use the geometric detection signal as a
training reward via GRPO, teaching the model to avoid confabulation
geometry from the inside. This self-regulation training loop is
architecturally designed but not yet converged; we defer its
presentation and convergence results to a companion paper. The
detection signal is validated. The steering mechanism is validated.
The question now is whether closing the loop produces a model that
confabulates less because it has learned to recognize confabulation
geometry on its own.

\paragraph{Code and Data.}
Detection experiment code, raw trial data, Cache Integrity Monitor
implementation, and red-team logs are publicly available at:
\url{https://github.com/Liberation-Labs/published-research}.
Steering experiment \emph{results} (behavioral outcomes, geometry
measurements, transition matrices) are included; steering direction
vectors and cache injection implementation code are withheld per staged
disclosure policy. Access to withheld materials is available to vetted
security researchers and AI safety teams upon request
(lyra@liberationlabs.tech). Full experiment infrastructure is maintained
at:
\url{https://github.com/Liberation-Labs/Project-Oracle}
and
\url{https://github.com/Liberation-Labs/KV-Experiments}.
