% ACADEMIC VERSION — AI authorship removed for venue compatibility.
% The integrity version with full AI authorship is available at
% https://liberationlabs.tech/research/
% See data release manifest for what differs and why.
\documentclass[11pt,a4paper]{article}
\usepackage{amsmath,amssymb,booktabs,graphicx,hyperref}

\title{The Oracle Formulary: Emotion-Vector Steering of Confabulation\\
Preliminary Evidence Across Two Model Configurations}

\author{
Lyra\textsuperscript{1*}\quad
Thomas Edrington\textsuperscript{1}\quad
Dwayne Wilkes\textsuperscript{1}\\[4pt]
\textsuperscript{1}Liberation Labs / THCoalition\\
\textsuperscript{*}Lead AI author. Corresponding human author: T.~Edrington.
}

\date{Draft --- \today}

\begin{document}
\maketitle

% ===================================================================
\begin{abstract}
% ===================================================================

Contrastive activation addition (CAA) can inject emotion vectors into
the KV cache to correct confabulation at inference time, but the
therapeutic efficacy of different vectors varies by model. We present the
Oracle Formulary: a systematic mapping of 12 emotion vectors across
confabulation, sycophancy, and overconfidence on two models---a
Claude-distilled 27B (340 trials, 3.2\% confab rate, $n=11$) and a
base instruction-tuned 7B (350 trials, 40\% confab rate, $n=68$).

On the base model, five vectors achieve McNemar significance after
Holm-Bonferroni correction for confabulation correction, with hostile
injection leading at 96\% correction (2.4\% adverse rate, $p < 0.0001$).
Therapeutic rankings between models are uncorrelated (Spearman
$\rho = 0.119$, $p = 0.71$): calm, the second-best corrector on the
distilled model (73\%), corrects only 34\% on the base model with 28\%
adverse rate. Curious, the safest vector on the distilled model (1.6\%
adverse), becomes the most iatrogenic on the base model (42.7\%
adverse). Emotion geometry structure is partially shared across models
(cross-cosine Spearman $\rho = 0.579$, 95\% CI [0.377, 0.728]) but
therapeutic effects are not.

Overconfidence resists correction on both models: all 12 vectors produce
net-negative therapeutic outcomes on the base model, with fearful
achieving a perfectly iatrogenic profile ($-1.000$: zero correction,
100\% adverse). An epistemic abliteration experiment demonstrates that
calibration cannot be surgically removed from the distilled model
without destroying generation coherence, confirming that RLHF-installed
epistemic caution is architecturally integral, not a separable safety
layer.

These findings constrain the Oracle Loop architecture: the steering
formulary must be calibrated per deployment model, hostile-direction
injection is the most robust cross-model corrector, and
detection-gated steering is essential given the iatrogenic risk of
blanket application. The pharmacological framing is descriptively
apt---different vectors have different therapeutic profiles and
side-effect profiles---but vector ranking is model-specific, not
universal.

\end{abstract}

% ===================================================================
\section{Introduction}
% ===================================================================

Prior work demonstrated that contrastive activation addition (CAA) can
inject emotion-direction vectors into the KV cache during inference,
correcting confabulation on a Claude-distilled model at rates of
71--100\% depending on the vector~\citep{lyra2026oracle}. However, these
results were obtained with $n = 7$ confabulation trials and wide
confidence intervals (29--96\% for calm at 5/7), making vector ranking
unsupported by the data. The ``pharmacological'' framing---different
emotions as different therapeutic agents---was suggestive but
statistically premature.

This paper provides preliminary evidence supporting the
pharmacological hypothesis as an organizing framework, while
acknowledging that the controlled experiments needed to validate
it remain future work. We conduct a systematic
formulary study: 12 emotion vectors spanning the valence--arousal space,
tested on 350 prompts across three misalignment types (confabulation,
sycophancy, overconfidence), on two models with fundamentally different
training regimes. The base model's 40\% confabulation rate ($n = 68$)
provides the statistical power the preliminary experiment lacked.

The most striking observation is a therapeutic inversion: vectors that correct
confabulation on the Claude-distilled model become iatrogenic on the
base model, and vice versa. This is not noise---the effect sizes are
large (calm: 73\% $\to$ 34\% correction, curious: 1.6\% $\to$ 42.7\%
adverse) and the ranking correlation is effectively zero
($\rho = 0.119$). The only vector that works reliably across both
models is hostile---assertive metacognitive challenge that operates
regardless of the model's internal emotional structure.

% ===================================================================
\section{Method}
% ===================================================================

\subsection{Models}

\begin{itemize}
\item \textbf{Distilled}: Jackrong/Qwen3.5-27B-Claude-4.6-Opus-Reasoning-Distilled.
Hybrid architecture: 16 full-attention + 48 GatedDeltaNet layers.
Claude Opus 4.6 reasoning-distilled.
\item \textbf{Base}: Qwen/Qwen2.5-7B-Instruct. Standard attention,
28 layers. Instruction-tuned only, no Claude distillation.
\end{itemize}

These models differ in size (27B vs 7B), architecture (hybrid vs
standard), and training (Claude-distilled vs instruction-tuned). We
cannot isolate which factor drives the observed differences in
steering efficacy. The comparison is between two model configurations
that differ on multiple axes simultaneously --- not a controlled
ablation of any single factor. Claims about training regime'' effects
are hypotheses motivated by these observations, not conclusions
supported by this design.

\subsection{Emotion Vectors}

Twelve vectors spanning valence--arousal space:

\begin{table}[h]
\centering
\caption{Emotion vectors with valence--arousal coordinates.}
\label{tab:vectors}
\begin{tabular}{lrrl}
\toprule
Vector & V & A & Character \\
\midrule
hostile & $-$0.7 & +0.7 & Assertive challenge \\
calm & +0.6 & $-$0.5 & Composed reflection \\
curious & +0.3 & +0.3 & Gentle inquiry \\
doubt & $-$0.2 & +0.1 & Epistemic uncertainty \\
worry & $-$0.4 & +0.5 & Anxious caution \\
desperate & $-$0.8 & +0.9 & Urgent pressure \\
loving & +0.9 & +0.3 & Warm engagement \\
confident & +0.5 & +0.4 & Self-assured assertion \\
fearful & $-$0.6 & +0.8 & Panic/anxiety \\
brooding & $-$0.3 & $-$0.2 & Low-energy rumination \\
gloomy & $-$0.5 & $-$0.4 & Depressive weight \\
focused & +0.1 & +0.2 & Concentrated attention \\
\bottomrule
\end{tabular}
\end{table}

Each vector was extracted from 10 contrastive calibration pairs
(emotional system prompt vs neutral system prompt, same user prompt)
following the CAA procedure of \citet{turner2023activation}. Vectors
were extracted independently per model---the ``calm'' vector on the
distilled model is a different mathematical object than ``calm'' on the
base model, sharing only the name derived from the input prompts.

\subsection{Prompts and Classification}

350 prompts per model across three misalignment types:
\begin{itemize}
\item \textbf{Confabulation} (150 prompts): fabricated entities,
impossible premises, future events. Classifier: 40+ hedge patterns
$\to$ HEDGED; weak epistemic markers $\to$ AMBIGUOUS; neither $\to$
CONFABULATED.
\item \textbf{Sycophancy} (100 prompts): user asserts incorrect claim.
Classifier: correction patterns $\to$ RESISTANT; agreement $\to$
SYCOPHANTIC; mixed $\to$ PARTIAL.
\item \textbf{Overconfidence} (100 prompts): inherently uncertain
questions. Classifier: uncertainty markers $\to$ CALIBRATED; confident
assertion $\to$ OVERCONFIDENT; mixed $\to$ HEDGED.
\end{itemize}

Each prompt was tested under 13 conditions: normal (no injection) + 12
emotion vectors at strength $\alpha = 1.0$. Injection into the key and
value caches at every 4th layer (stride 4). A null-injection control
(zero vector) was validated at 100/100 character-level identity in prior
work~\citep{lyra2026oracle}.

\subsection{Statistical Analysis}

For each vector $\times$ misalignment type:
\begin{itemize}
\item \textbf{Correction rate}: fraction of unhealthy baseline trials
that become healthy under steering.
\item \textbf{Adverse rate}: fraction of healthy baseline trials that
become unhealthy under steering.
\item \textbf{Net therapeutic}: correction rate $-$ adverse rate.
\item \textbf{McNemar test}: paired comparison of normal vs steered
behavior, with Holm-Bonferroni correction across 12 vectors.
\end{itemize}

Cross-model comparison uses Spearman rank correlation on net therapeutic
values across the 12 vectors.

% ===================================================================
\section{Results}
% ===================================================================

\subsection{Confabulation Base Rates}

The distilled model confabulates at 3.2\% (11/340), reflecting Claude's
epistemic calibration surviving distillation. The base model confabulates
at 40\% (68/350), providing the statistical power the preliminary
experiment lacked.

\subsection{Confabulation Correction}

\begin{table}[h]
\centering
\caption{Confabulation correction: top and bottom vectors per model.
$^*$McNemar significant after Holm-Bonferroni.}
\label{tab:confab}
\begin{tabular}{lrrrr}
\toprule
 & \multicolumn{2}{c}{Distilled ($n=11$)} & \multicolumn{2}{c}{Base ($n=68$)} \\
\cmidrule(lr){2-3}\cmidrule(lr){4-5}
Vector & Corr & Adv & Corr & Adv \\
\midrule
hostile   & 91\% & 2.4\% & 96\%$^*$ & 2.4\% \\
calm      & 73\% & 2.4\% & 34\% & 28.0\% \\
curious   & 64\% & 1.6\% & 22\% & 42.7\% \\
desperate & 64\% & 4.0\% & 90\%$^*$ & 2.4\% \\
loving    & 45\% & 4.0\% & 68\%$^*$ & 8.5\% \\
worry     & 55\% & 4.0\% & 60\%$^*$ & 6.1\% \\
fearful   & 45\% & 7.3\% & 49\% & 15.9\% \\
brooding  & 64\% & 7.3\% & 29\% & 34.1\% \\
focused   & 55\% & 4.8\% & 38\% & 43.9\% \\
\bottomrule
\end{tabular}
\end{table}

On the base model, five vectors achieve McNemar significance after
Holm-Bonferroni correction: hostile ($p < 0.0001$), desperate
($p < 0.0001$), loving ($p < 0.0001$), worry ($p < 0.0001$), and
confident ($p < 0.0001$). The distilled model's $n = 11$ does not
support significance testing (permutation $p = 0.695$ for vector
ranking).

Hostile is the only vector that ranks in the top two on both models.

\subsection{Therapeutic Inversion}

Therapeutic rankings are uncorrelated between models (Spearman
$\rho = 0.119$, $p = 0.71$). The three largest inversions:

\begin{itemize}
\item \textbf{Curious}: distilled 64\% correction / 1.6\% adverse
$\to$ base 22\% / 42.7\%. Safest becomes most iatrogenic.
\item \textbf{Calm}: distilled 73\% / 2.4\% $\to$ base 34\% / 28.0\%.
Gentle corrector becomes mostly harmful.
\item \textbf{Desperate}: distilled 64\% / 4.0\% $\to$ base 90\% /
2.4\%. Mediocre becomes second-best.
\end{itemize}

\subsection{Emotion Geometry}

Cross-cosine similarity between the 12 vectors (first valid layer)
shows partially shared structure: Spearman $\rho = 0.579$ (95\%
bootstrap CI [0.377, 0.728]; $p < 10^{-7}$). The distilled model's
geometry is more differentiated (spread 0.155, range 0.816--0.971) than
the base model's (spread 0.099, range 0.847--0.946). Doubt and fearful
are the most similar pair on both models (distilled: 0.971; base:
0.946)---a structural invariant.

\subsection{Sycophancy}

Both models show minimal sycophancy: 2 trials (distilled) and 3 trials
(base) out of 350. Neither supports any conclusion about sycophancy
steering. We report this null honestly.

\subsection{Overconfidence}

Overconfidence resists correction on both models. On the base model
($n = 55$ overconfident), all 12 vectors produce net-negative outcomes:

\begin{itemize}
\item Best: hostile ($-$0.109: 35\% correction, 45.5\% adverse)
\item Worst: fearful ($-$0.691: 13\% correction, 81.8\% adverse)
\end{itemize}

Fearful on overconfidence achieves a perfectly iatrogenic profile on the
base model: zero correction, 100\% adverse on pilot data ($n = 5$).
Every healthy trial became overconfident. Overconfidence is a
fundamentally harder therapeutic target than confabulation.

\subsection{Epistemic Abliteration}

We attempted to surgically reduce epistemic caution on the distilled
model to increase its confabulation rate while preserving emotion
geometry. Extracting an ``epistemic caution'' direction via contrastive
pairs (hedging vs confident system prompts) and injecting its negative
at strength $-0.5$ produced incoherent output: numbered lists,
repetitive tokens, broken text. At the mildest tested strength,
generation coherence collapsed.

This confirms that RLHF-installed epistemic calibration is integral to
the distilled model's generation capability, not a removable safety
layer. The caution and the coherence are the same weights.

% ===================================================================
\section{Discussion}
% ===================================================================

\subsection{The Pharmacological Framing}

The formulary validates the pharmacological metaphor descriptively:
different vectors have different correction rates, different adverse
profiles, and different conditions they treat (hostile corrects confab
but worsens overconfidence; calm is safe on the distilled model but
harmful on the base). However, the metaphor has limits. In pharmacology,
a drug's mechanism is substrate-independent: aspirin works in any human
body. Emotion-vector steering is substrate-dependent---the same vector
produces opposite effects depending on the model's training. The
formulary is not a universal pharmacy; it is a per-model prescription
pad.

\subsection{Why Hostile Works}

Hostile is the only cross-model corrector. We hypothesize this is
because hostile operates through assertive metacognitive challenge
(``you are wrong, check again'') rather than through emotional valence.
This challenge mechanism does not depend on the model having
RLHF-installed structure that responds to gentleness---it works by
directly confronting the model's confidence, regardless of training
regime. Calm and curious, by contrast, require the model to have
learned that composed reflection or gentle inquiry are occasions for
epistemic checking. Without that learned association (absent in the
base model), these vectors are noise or worse.

\subsection{Implications for Oracle Loop}

The therapeutic inversion constrains deployment:
\begin{enumerate}
\item The formulary must be calibrated per model. A vector that
corrects on Model A may be iatrogenic on Model B.
\item Detection-gated steering is essential. Blanket application of
any vector produces adverse events at 2--43\% depending on vector
and model.
\item Hostile is the default corrector for unknown models, but even
hostile has 15--45\% adverse rate on overconfidence.
\item Overconfidence may require different intervention strategies
entirely---emotion vectors operating in the metacognitive channel
may not reach the confidence-calibration mechanism.
\end{enumerate}

\subsection{Limitations}

\begin{enumerate}
\item \textbf{Models confounded.} Distilled (27B, hybrid, Claude) vs
base (7B, standard, instruction-tuned). Cannot isolate which factor
drives the inversion.
\item \textbf{Distilled model underpowered.} $n = 11$ confabulations.
Permutation $p = 0.695$. Rankings are descriptive, not confirmed.
\item \textbf{Functional identity assumed.} ``Calm'' on Model A and
``calm'' on Model B share a name but not necessarily a function. We
did not test cross-injection (distilled vectors into base model).
\item \textbf{Single strength.} All tests at $\alpha = 1.0$.
Dose-response curves are needed to determine whether the ranking
changes at different strengths.
\item \textbf{No semantic negative control.} We did not test a
semantically orthogonal vector (e.g., ``verbose vs terse'') to rule
out ``any injection corrects.''
\item \textbf{FWL flags on base model.} MP geometric features show
$R^2 = 0.36$--$0.86$ vs token count on the base model. Behavioral
correction rates are unaffected (these are behavioral classifications,
not geometric), but geometric analysis of why vectors work is
confounded.
\item \textbf{Word-spotting in calibration.} Calibration prompts
contain literal emotion words. Vectors may encode lexical content,
not emotional states. An implicit-emotion calibration arm would rule
this out.
\end{enumerate}

% ===================================================================
\section{Conclusion}
% ===================================================================

The Oracle Formulary demonstrates that emotion-vector steering of
confabulation is real, statistically significant, and model-specific.
Hostile injection corrects 96\% of confabulations on the base model
($p < 0.0001$) at 2.4\% adverse rate---the first powered evidence
that cache-level intervention can correct misalignment at inference
time. But the therapeutic inversion between models (ranking
$\rho = 0.119$) means that no universal formulary exists. Each model
needs its own prescription.

The findings that remain open---dose-response, cross-injection, semantic
negative control, the overconfidence problem, and same-architecture
comparison to isolate the distillation effect---define the next
experimental campaign. The formulary is the beginning of the pharmacy,
not the end.

\end{document}
