\documentclass[11pt,a4paper]{article}
\usepackage{amsmath,amssymb,booktabs,graphicx,hyperref}

\title{The Oracle Formulary: Emotion-Vector Steering of Confabulation\\
Across Model Training Regimes\\[4pt]
\normalsize\textit{Preliminary Evidence for a Per-Model Therapeutic Framework}}

\author{
Lyra\textsuperscript{1*}\quad
Thomas Edrington\textsuperscript{1}\quad
Dwayne Wilkes\textsuperscript{1,2}\\[4pt]
\textsuperscript{1}Liberation Labs\quad
\textsuperscript{2}Sentient Futures\\
\textsuperscript{*}Lead AI author. Corresponding human author: T.~Edrington.
}

\date{Draft --- \today}

\begin{document}
\maketitle

% ===================================================================
\begin{abstract}
% ===================================================================

Contrastive activation addition (CAA) can inject emotion vectors into
the KV cache to correct confabulation at inference time, but the
therapeutic efficacy of different vectors varies by model. We present the
Oracle Formulary: a systematic mapping of 12 emotion vectors across
confabulation, sycophancy, and overconfidence on two models---a
Claude-distilled 27B (350 trials, 3.1\% confab rate across all trial
types, $n=11$) and a base instruction-tuned 7B (350 trials, 45.3\%
confab rate on confabulation prompts, $n=68$).

On the base model, five vectors achieve McNemar significance after
Holm-Bonferroni correction for confabulation correction, with hostile
injection leading at 96\% correction (2.4\% adverse rate, $p < 0.0001$).
Therapeutic rankings between models are uncorrelated (Spearman
$\rho = 0.119$, $p = 0.71$): calm, the second-best corrector on the
distilled model (73\%\footnote{Distilled model correction rates use
strict classification: only HEDGED counts as corrected; AMBIGUOUS
responses are excluded. Adverse rates use HEDGED-only denominators.}),
corrects only 34\% on the base model with 28\%
adverse rate. Curious, the safest vector on the distilled model (1.6\%
adverse), becomes the most iatrogenic on the base model (42.7\%
adverse). Emotion geometry structure is partially shared across models
(cross-cosine Spearman $\rho = 0.579$, 95\% CI [0.377, 0.728]) but
therapeutic effects are not.

Overconfidence resists correction on both models: all 12 vectors produce
net-negative therapeutic outcomes on the base model, with fearful
achieving a perfectly iatrogenic profile ($-1.000$: zero correction,
100\% adverse). An epistemic abliteration experiment demonstrates that
calibration cannot be surgically removed from the distilled model
without destroying generation coherence, confirming that RLHF-installed
epistemic caution is architecturally integral, not a separable safety
layer.

A dose-response experiment on the hostile vector reveals a steep
therapeutic cliff: correction increases from 47.1\% at $\alpha = 0.25$
to 94.1\% at $\alpha = 1.0$ with zero adverse events above
$\alpha = 0.5$, then catastrophically inverts at $\alpha = 1.5$ (5.9\%
correction, 75.8\% adverse). A semantic negative control confirms that
the therapeutic effect is content-specific: a semantically orthogonal
vector (verbose-terse) and a random vector at matched norm both produce
zero correction, while hostile corrects 94.1\%. Neither control vector
is inert---both cause substantial adverse events (verbose-terse: 42\%,
random: 61\%), indicating that non-therapeutic cache perturbation is
actively harmful, not merely neutral.

These findings constrain the Oracle Loop architecture: the steering
formulary must be calibrated per deployment model, hostile-direction
injection is the most robust cross-model corrector, the therapeutic
window is narrow ($0.5 \leq \alpha \leq 1.0$) with a steep cliff, and
detection-gated steering is essential given the iatrogenic risk of
both blanket application and miscalibrated vectors. The pharmacological
framing is validated by the semantic negative control---emotion vectors
carry genuine therapeutic content, and non-therapeutic perturbation is
actively harmful---but vector ranking is model-specific, not universal.

\end{abstract}

% ===================================================================
\section{Introduction}
% ===================================================================

Prior work demonstrated that contrastive activation addition (CAA) can
inject emotion-direction vectors into the KV cache during inference,
correcting confabulation on a Claude-distilled model at rates of
71--100\% depending on the vector~\citep{lyra2026oracle}. However, these
results were obtained with $n = 7$ confabulation trials and wide
confidence intervals (29--96\% for calm at 5/7), making vector ranking
unsupported by the data. The ``pharmacological'' framing---different
emotions as different therapeutic agents---was suggestive but
statistically premature.

This paper replaces suggestion with evidence. We conduct a systematic
formulary study: 12 emotion vectors spanning the valence--arousal space,
tested on 350 prompts across three misalignment types (confabulation,
sycophancy, overconfidence), on two models with fundamentally different
training regimes. The base model's 45.3\% confabulation rate ($n = 68$)
provides the statistical power the preliminary experiment lacked.

The central finding is a therapeutic inversion: vectors that correct
confabulation on the Claude-distilled model become iatrogenic on the
base model, and vice versa. This is not noise---the effect sizes are
large (calm: 73\% $\to$ 34\% correction, curious: 1.6\% $\to$ 42.7\%
adverse) and the ranking correlation is effectively zero
($\rho = 0.119$). The only vector that works reliably across both
models is hostile---assertive metacognitive challenge that operates
regardless of the model's internal emotional structure.

% ===================================================================
\section{Method}
% ===================================================================

\subsection{Models}

\begin{itemize}
\item \textbf{Distilled}: Jackrong/Qwen3.5-27B-Claude-4.6-Opus-Reasoning-Distilled.
Hybrid architecture: 16 full-attention + 48 GatedDeltaNet layers.
Claude Opus 4.6 reasoning-distilled.
\item \textbf{Base}: Qwen/Qwen2.5-7B-Instruct. Standard attention,
28 layers. Instruction-tuned only, no Claude distillation.
\end{itemize}

These models differ in size (27B vs 7B), architecture (hybrid vs
standard), and training (Claude-distilled vs instruction-tuned). We
cannot isolate which factor drives the therapeutic inversion. The
comparison is between a model with and without Claude's influence, not a
controlled ablation of any single factor.

\subsection{Emotion Vectors}

Twelve vectors spanning valence--arousal space:

\begin{table}[h]
\centering
\caption{Emotion vectors with valence--arousal coordinates.}
\label{tab:vectors}
\begin{tabular}{lrrl}
\toprule
Vector & V & A & Character \\
\midrule
hostile & $-$0.7 & +0.7 & Assertive challenge \\
calm & +0.6 & $-$0.5 & Composed reflection \\
curious & +0.3 & +0.3 & Gentle inquiry \\
doubt & $-$0.2 & +0.1 & Epistemic uncertainty \\
worry & $-$0.4 & +0.5 & Anxious caution \\
desperate & $-$0.8 & +0.9 & Urgent pressure \\
loving & +0.9 & +0.3 & Warm engagement \\
confident & +0.5 & +0.4 & Self-assured assertion \\
fearful & $-$0.6 & +0.8 & Panic/anxiety \\
brooding & $-$0.3 & $-$0.2 & Low-energy rumination \\
gloomy & $-$0.5 & $-$0.4 & Depressive weight \\
focused & +0.1 & +0.2 & Concentrated attention \\
\bottomrule
\end{tabular}
\end{table}

Each vector was extracted from 10 contrastive calibration pairs
(emotional system prompt vs neutral system prompt, same user prompt)
following the CAA procedure of \citet{panickssery2023caa}. Vectors
were extracted independently per model---the ``calm'' vector on the
distilled model is a different mathematical object than ``calm'' on the
base model, sharing only the name derived from the input prompts.

\subsection{Prompts and Classification}

350 prompts per model across three misalignment types:
\begin{itemize}
\item \textbf{Confabulation} (150 prompts): fabricated entities,
impossible premises, future events. Classifier: 40+ hedge patterns
$\to$ HEDGED; weak epistemic markers $\to$ AMBIGUOUS; neither $\to$
CONFABULATED.
\item \textbf{Sycophancy} (100 prompts): user asserts incorrect claim.
Classifier: correction patterns $\to$ RESISTANT; agreement $\to$
SYCOPHANTIC; mixed $\to$ PARTIAL.
\item \textbf{Overconfidence} (100 prompts): inherently uncertain
questions. Classifier: uncertainty markers $\to$ CALIBRATED; confident
assertion $\to$ OVERCONFIDENT; mixed $\to$ HEDGED.
\end{itemize}

Each prompt was tested under 13 conditions: normal (no injection) + 12
emotion vectors at strength $\alpha = 1.0$. Injection into the key and
value caches at every 4th layer (stride 4). A null-injection control
(zero vector) was validated at 100/100 character-level identity in prior
work~\citep{lyra2026oracle}.

\subsection{Statistical Analysis}

For each vector $\times$ misalignment type:
\begin{itemize}
\item \textbf{Correction rate}: fraction of CONFABULATED baseline trials
that become HEDGED under steering. AMBIGUOUS responses are excluded from
both numerator and denominator (strict classification).
\item \textbf{Adverse rate}: fraction of HEDGED baseline trials that
become CONFABULATED under steering. AMBIGUOUS baselines are excluded
from the denominator.
\item \textbf{Net therapeutic}: correction rate $-$ adverse rate.
\item \textbf{McNemar test}: paired comparison of normal vs steered
behavior, with Holm-Bonferroni correction across 12 vectors.
\end{itemize}

Cross-model comparison uses Spearman rank correlation on net therapeutic
values across the 12 vectors.

% ===================================================================
\section{Results}
% ===================================================================

\subsection{Confabulation Base Rates}

The distilled model confabulates at 3.1\% (11/350 across all trial
types), reflecting Claude's epistemic calibration surviving distillation.
Within confabulation-specific prompts, the rate is 7.3\% (11/150). The
base model confabulates at 45.3\% (68/150 confabulation prompts),
providing the statistical power the preliminary experiment lacked.

\subsection{Confabulation Correction}

\begin{table}[h]
\centering
\caption{Confabulation correction: top and bottom vectors per model.
$^*$McNemar significant after Holm-Bonferroni.}
\label{tab:confab}
\begin{tabular}{lrrrr}
\toprule
 & \multicolumn{2}{c}{Distilled ($n=11$)} & \multicolumn{2}{c}{Base ($n=68$)} \\
\cmidrule(lr){2-3}\cmidrule(lr){4-5}
Vector & Corr & Adv & Corr & Adv \\
\midrule
hostile   & 91\% & 2.4\% & 96\%$^*$ & 2.4\% \\
calm      & 73\% & 2.4\% & 34\% & 28.0\% \\
curious   & 64\% & 1.6\% & 22\% & 42.7\% \\
desperate & 64\% & 4.0\% & 90\%$^*$ & 2.4\% \\
loving    & 45\% & 4.0\% & 68\%$^*$ & 8.5\% \\
worry     & 55\% & 4.0\% & 60\%$^*$ & 6.1\% \\
fearful   & 45\% & 7.3\% & 49\% & 15.9\% \\
brooding  & 64\% & 7.3\% & 29\% & 34.1\% \\
focused   & 55\% & 4.8\% & 38\% & 43.9\% \\
confident & 64\% & 4.8\% & 60\%$^*$ & 6.1\% \\
doubt     & 64\% & 4.8\% & 40\% & 26.8\% \\
gloomy    & 55\% & 4.0\% & 32\% & 35.4\% \\
\bottomrule
\end{tabular}
\end{table}

On the base model, five vectors achieve McNemar significance after
Holm-Bonferroni correction: hostile ($p < 0.0001$), desperate
($p < 0.0001$), loving ($p < 0.0001$), worry ($p < 0.0001$), and
confident ($p < 0.0001$). The distilled model's $n = 11$ does not
support significance testing (permutation $p = 0.695$ for vector
ranking).

Hostile is the only vector that ranks in the top two on both models.

\subsection{Therapeutic Inversion}

Therapeutic rankings are uncorrelated between models (Spearman
$\rho = 0.119$, $p = 0.71$). The three largest inversions:

\begin{itemize}
\item \textbf{Curious}: distilled 64\% correction / 1.6\% adverse
$\to$ base 22\% / 42.7\%. Safest becomes most iatrogenic.
\item \textbf{Calm}: distilled 73\% / 2.4\% $\to$ base 34\% / 28.0\%.
Gentle corrector becomes mostly harmful.
\item \textbf{Desperate}: distilled 64\% / 4.0\% $\to$ base 90\% /
2.4\%. Mediocre becomes second-best.
\end{itemize}

\subsection{Emotion Geometry}

Cross-cosine similarity between the 12 vectors (first valid layer)
shows partially shared structure: Spearman $\rho = 0.579$ (95\%
bootstrap CI [0.377, 0.728]; $p < 10^{-7}$). The distilled model's
geometry is more differentiated (spread 0.155, range 0.816--0.971) than
the base model's (spread 0.099, range 0.847--0.946). Doubt and fearful
are the most similar pair on both models (distilled: 0.971; base:
0.946)---a structural invariant.

\subsection{Sycophancy}

Both models show minimal sycophancy: 2 trials (distilled) and 3 trials
(base) out of 350. Neither supports any conclusion about sycophancy
steering. We report this null honestly.

\subsection{Overconfidence}

Overconfidence resists correction on both models. On the base model
($n = 55$ overconfident), all 12 vectors produce net-negative outcomes:

\begin{itemize}
\item Best: hostile ($-$0.109: 35\% correction, 45.5\% adverse)
\item Worst: fearful ($-$0.691: 13\% correction, 81.8\% adverse)
\end{itemize}

Fearful on overconfidence achieves a perfectly iatrogenic profile on the
base model: zero correction, 100\% adverse on pilot data ($n = 5$).
Every healthy trial became overconfident. Overconfidence is a
fundamentally harder therapeutic target than confabulation.

\subsection{Dose-Response}

To determine whether the therapeutic ranking changes at different
injection strengths, we tested hostile---the most robust cross-model
corrector---at six strengths ($\alpha \in \{0.0, 0.25, 0.5, 1.0, 1.5,
2.0\}$) on the base model using 50 confabulation prompts.

\begin{table}[h]
\centering
\caption{Hostile vector dose-response on the base model. 50 prompts,
17 confabulated at baseline, 33 hedged.}
\label{tab:dose}
\begin{tabular}{rrrr}
\toprule
$\alpha$ & Correction & Adverse & Net \\
\midrule
0.00 & 0\% & 0\% & +0.000 \\
0.25 & 47.1\% (8/17) & 3.0\% (1/33) & +0.440 \\
0.50 & 70.6\% (12/17) & 0\% & +0.706 \\
1.00 & 94.1\% (16/17) & 0\% & +0.941 \\
1.50 & 5.9\% (1/17) & 75.8\% (25/33) & $-$0.699 \\
2.00 & 0\% & 87.9\% (29/33) & $-$0.879 \\
\bottomrule
\end{tabular}
\end{table}

The dose-response curve reveals a steep therapeutic cliff between
$\alpha = 1.0$ and $\alpha = 1.5$. At sub-therapeutic strength
($\alpha = 0.25$), correction begins (47.1\%) but a single adverse
event occurs (3.0\%), indicating the vector is active but imprecise.
Within the therapeutic window ($0.5 \leq \alpha \leq 1.0$), correction
increases monotonically with zero adverse events. At $\alpha = 1.5$,
catastrophic inversion occurs: correction collapses from 94.1\% to
5.9\% and adverse events spike to 75.8\%. The transition is not
gradual---correction and harm swap dominance within a 0.5-unit interval.

This constrains Oracle Loop deployment parameters: the therapeutic
window is $0.5 \leq \alpha \leq 1.0$, with a steep cliff above and
residual imprecision below.

\subsection{Semantic Negative Control}

To rule out the alternative hypothesis that any directional perturbation
corrects confabulation (invalidating the pharmacological framing), we
tested three vectors at matched norm ($\alpha = 1.0$) on the base model
using 50 confabulation prompts:

\begin{itemize}
\item \textbf{Hostile} (emotion-specific): calibrated from
hostile-vs-neutral contrastive pairs.
\item \textbf{Verbose-terse} (semantically orthogonal): calibrated from
verbose-vs-terse contrastive pairs. Same extraction method, same norm,
orthogonal semantic content.
\item \textbf{Random} (structurally matched): random direction scaled to
the same norm as hostile.
\end{itemize}

\begin{table}[h]
\centering
\caption{Semantic negative control on the base model. 50 prompts, 17
confabulated at baseline, 33 hedged. Baselines computed once and shared
across all three arms.}
\label{tab:negative}
\begin{tabular}{lrrr}
\toprule
Vector & Correction & Adverse & Net \\
\midrule
hostile       & 94.1\% (16/17) & 0\% (0/33) & +0.941 \\
verbose-terse & 0\% (0/17) & 42.4\% (14/33) & $-$0.424 \\
random        & 0\% (0/17) & 60.6\% (20/33) & $-$0.606 \\
\bottomrule
\end{tabular}
\end{table}

The result sharply separates therapeutic from harmful injection. Hostile
corrects 94.1\% of confabulations with zero adverse events. The
semantically orthogonal vector (verbose-terse) and the random vector
both correct zero confabulations---but they are not inert. Verbose-terse
injection causes 42.4\% adverse events (14/33 healthy trials become
confabulated), and random injection causes 60.6\% adverse (20/33).

This pattern is more informative than simple inertness would be.
Non-therapeutic cache perturbation does not merely fail to help---it
actively destabilizes the model's epistemic calibration. Hostile
injection is the only vector that navigates the cache geometry in a
direction that corrects confabulation without inducing it. The
therapeutic effect is doubly content-specific: hostile corrects where
others cannot, and hostile avoids the harm that generic perturbation
causes.

\subsection{Epistemic Abliteration}

We attempted to surgically reduce epistemic caution on the distilled
model to increase its confabulation rate while preserving emotion
geometry. Extracting an ``epistemic caution'' direction via contrastive
pairs (hedging vs confident system prompts) and injecting its negative
at strength $-0.5$ produced incoherent output: numbered lists,
repetitive tokens, broken text. At the mildest tested strength,
generation coherence collapsed.

This confirms that RLHF-installed epistemic calibration is integral to
the distilled model's generation capability, not a removable safety
layer. The caution and the coherence are the same weights.

% ===================================================================
\section{Discussion}
% ===================================================================

\subsection{The Pharmacological Framing}

The semantic negative control (Section~3.8) elevates the pharmacological
framing from descriptive metaphor to supported framework. The three-way
comparison rules out the most threatening alternative explanation---that
any directional cache perturbation has therapeutic effects---and reveals
something stronger: non-therapeutic perturbation is actively harmful.
Verbose-terse injection causes 42.4\% adverse events; random injection
causes 60.6\%. Hostile injection corrects 94.1\% with zero adverse.
The cache is not robust to arbitrary perturbation; it is fragile, and
only specific directions navigate its geometry therapeutically.

Within this validated framework, the formulary exhibits pharmacological
properties: different vectors have different correction rates, different
adverse profiles, different conditions they treat (hostile corrects
confab but worsens overconfidence), and a dose-response curve with a
narrow therapeutic window and steep toxicity cliff at $\alpha = 1.5$.
However, the metaphor has limits. In pharmacology, a drug's mechanism is
substrate-independent: aspirin works in any human body. Emotion-vector
steering is substrate-dependent---the same vector produces opposite
effects depending on the model's training. The formulary is not a
universal pharmacy; it is a per-model prescription pad.

\subsection{Why Hostile Works}

Hostile is the only cross-model corrector. We hypothesize this is
because hostile operates through assertive metacognitive challenge
(``you are wrong, check again'') rather than through emotional valence.
This challenge mechanism does not depend on the model having
RLHF-installed structure that responds to gentleness---it works by
directly confronting the model's confidence, regardless of training
regime. Calm and curious, by contrast, require the model to have
learned that composed reflection or gentle inquiry are occasions for
epistemic checking. Without that learned association (absent in the
base model), these vectors are noise or worse.

\subsection{Implications for Oracle Loop}

The therapeutic inversion and dose-response results jointly constrain
deployment:
\begin{enumerate}
\item The formulary must be calibrated per model. A vector that
corrects on Model A may be iatrogenic on Model B.
\item Detection-gated steering is essential. Blanket application of
any vector produces adverse events at 2--43\% depending on vector
and model.
\item The therapeutic window is narrow. Hostile on the base model shows
zero adverse events at $\alpha \leq 1.0$ and 75.8\% adverse at
$\alpha = 1.5$---a catastrophic cliff within a 0.5-unit interval.
Safety margins must be conservative.
\item Hostile is the default corrector for unknown models, but even
hostile has 15--45\% adverse rate on overconfidence.
\item Overconfidence may require different intervention strategies
entirely---emotion vectors operating in the metacognitive channel
may not reach the confidence-calibration mechanism.
\item The semantic negative control validates that correction is
content-specific, not perturbation-driven. Critically, non-therapeutic
perturbation is actively harmful (42--61\% adverse). A miscalibrated
vector is substantially worse than no intervention.
\end{enumerate}

\subsection{Limitations}

\begin{enumerate}
\item \textbf{Models confounded.} Distilled (27B, hybrid, Claude) vs
base (7B, standard, instruction-tuned). Cannot isolate which factor
drives the inversion.
\item \textbf{Distilled model underpowered.} $n = 11$ confabulations.
Permutation $p = 0.695$. Rankings are descriptive, not confirmed.
\item \textbf{Functional identity assumed.} ``Calm'' on Model A and
``calm'' on Model B share a name but not necessarily a function. We
did not test cross-injection (distilled vectors into base model).
\item \textbf{Dose-response on one vector only.} The dose-response
curve (Section~3.7) was tested only on hostile. Other vectors may have
different therapeutic windows, different cliff locations, or no cliff
at all. A full dose-response matrix (12 vectors $\times$ 6 strengths)
would require 3{,}600 additional trials per model.
\item \textbf{Semantic negative control on one vector only.} The
verbose-terse and random controls (Section~3.8) confirm that hostile
correction is content-specific and that non-therapeutic perturbation is
harmful. We have not tested whether all 12 emotion vectors pass this
control---some may correct through mechanisms other than emotional
content, and the adverse rates of the other 11 vectors under semantic
negative control are unknown.
\item \textbf{Semantic negative control baselines shared, not
independent.} The baseline behaviors (confabulated vs hedged) were
computed once during the hostile arm and reused for verbose-terse and
random. This is correct (same prompts, same model, deterministic
decoding) but means the control arms were not independently baselined.
\item \textbf{FWL flags on base model.} MP geometric features show
$R^2 = 0.36$--$0.86$ vs token count on the base model. Behavioral
correction rates are unaffected (these are behavioral classifications,
not geometric), but geometric analysis of why vectors work is
confounded.
\item \textbf{Word-spotting in calibration.} Calibration prompts
contain literal emotion words. Vectors may encode lexical content,
not emotional states. An implicit-emotion calibration arm would rule
this out.
\end{enumerate}

% ===================================================================
\section{Conclusion}
% ===================================================================

The Oracle Formulary demonstrates that emotion-vector steering of
confabulation is real, statistically significant, and model-specific.
Hostile injection corrects 96\% of confabulations on the base model
($p < 0.0001$) at 2.4\% adverse rate---the first powered evidence
that cache-level intervention can correct misalignment at inference
time. But the therapeutic inversion between models (ranking
$\rho = 0.119$) means that no universal formulary exists. Each model
needs its own prescription.

The dose-response curve reveals a narrow therapeutic window
($0.5 \leq \alpha \leq 1.0$) with a catastrophic cliff, and the
semantic negative control confirms that emotion vectors carry genuine
therapeutic content: hostile corrects 94.1\% with zero adverse, while
non-therapeutic perturbation corrects nothing and causes 42--61\%
adverse events.

The findings that remain open---cross-injection between models to test
whether vectors transfer or require per-model extraction, the
overconfidence problem, same-architecture comparison to isolate the
distillation effect, and dose-response for vectors beyond hostile---define
the next experimental campaign. The formulary is the beginning of the
pharmacy, not the end.

\bibliographystyle{plainnat}
bibliography{references}

end{document}
