We present a method for detecting and correcting confabulation in language
models at inference time, requiring no access to model weights, no training
data, and no modification to the model itself. Singular value decomposition
(SVD) of the key-value cache---a standard byproduct of autoregressive
generation---produces spectral features that distinguish genuine reasoning
from confabulation. We introduce Marchenko-Pastur (MP) corrected features
derived from random matrix theory that are approximately
dimension-invariant, eliminating the dominant token-count confound. In a
detection experiment across three architectures (Qwen2.5-7B, Llama-3.1-8B,
Mistral-7B-v0.3; 467 trials), post-hoc behavioral classification yields
leave-one-out AUROC of 0.707 (Qwen, $p=0.001$) using MP features alone,
with encoding-phase features at chance (0.511), confirming a
generation-specific signal. In a separate steering experiment on
Qwen3.5-27B-Claude-Opus-Distilled (100 trials, five-arm within-subject
including no-op null verification),
calm-direction injection into the KV cache corrected 5 of 7 confabulations
while a null injection control corrected 0 of 7 ($p=0.031$, one-sided;
$p=0.063$, two-sided), with the null control producing
identical output on all 100 trials. An expanded five-arm experiment
(Normal, Null, Calm, Worry, Doubt) confirms a pharmacological pattern:
doubt corrects 7/7 (100\%), worry 6/7 (86\%), calm 5/7 (71\%), while
zero non-confabulating trials resist all three treatments. However,
blanket steering also induced adverse confabulation in 3--5\% of hedged
trials, confirming that detection-gated correction is required. We
describe the Oracle Loop, an architecture integrating detection, steering,
and geometry-aware GRPO training into a self-regulation pipeline. Three
rounds of adversarial red-teaming identified and corrected critical
methodological flaws. Recognizing that the steering methodology creates
a dual-use attack surface, we present a Cache Integrity Monitor that
detects unauthorized cache modifications against uniform additive
perturbation on two architectures (standard transformer and hybrid
attention) with 0/36 false positives (95\% CI: 0--10\%) and 72/72
true positives (95\% CI: 95--100\%) at injection strengths as low as
1\% of the operational level. Detection code, data, defense implementation, and red-team logs are
publicly available; steering vectors are available to vetted researchers
under staged disclosure.
