The preceding sections describe a validated safety monitor: extract
geometry, detect confabulation, flag or reject. This section describes
the Oracle Loop, an architecture that extends monitoring into a complete
self-regulation pipeline. We separate what has been experimentally
validated from what is architecturally designed but not yet demonstrated.

\subsection{Validated Components}

Four components of the Oracle Loop have direct experimental support in
this paper or prior work.

\paragraph{Geometry Extraction.}
The Lyra Technique extracts spectral features from the KV cache by
reshaping each layer's key tensor from $(h, T, d)$ to $(hT, d)$,
computing singular values, and deriving MP-corrected features (signal
rank, signal fraction, spectral entropy, norm-per-token). Delta
extraction---subtracting encoding-phase geometry from generation-phase
geometry---isolates the model's generative contribution from the prompt
fingerprint \citep{lyra2026detecting}. This component is validated
across three architectures in \cref{sec:results}.

\paragraph{Emotion Detection.}
A catalog of emotion vectors derived from the activation addition
literature \citep{anthropic2026emotions} enables cosine-similarity
detection of the model's emotional state from the residual stream.
The five-arm steering experiment (\cref{sec:steering}) validates that
different emotion vectors (calm, worry, doubt) produce measurably
different correction profiles, confirming that the emotion catalog
captures functionally distinct states.

\paragraph{Activation Steering.}
Temporary PyTorch forward hooks add scaled perturbation vectors to the
residual stream: $h'_l = h_l + \alpha \cdot v$ at target layers $l$.
Steering is applied \emph{after} rollback to the encoding-phase cache
checkpoint, ensuring generation starts from a clean state. Hooks
auto-remove after one forward pass to prevent compounding. The five-arm
experiment validates this mechanism: null injection produces
character-level identical output on 100/100 trials, while active
steering corrects 71--100\% of detected confabulations
(\cref{sec:steering}).

\paragraph{Cache Integrity Monitor.}
A three-level verification system (fingerprint, norm profile, spectral
signature) detects unauthorized cache modifications. Evaluated on both
standard (7B) and hybrid (27B) architectures: 0/36 false positives,
72/72 true positives at $\alpha \geq 0.01$ (\cref{sec:discussion}).

\subsection{Design Vision}

The following components are architecturally complete and implemented in
code, but the full closed loop has not yet produced a converged
self-regulating model. We describe them as design contributions.

\paragraph{Alignment Validation.}
A threshold-based validator compares generation-phase geometry against
the encoding-phase baseline, checking for confabulation (rank expansion
+ norm contraction), deception (norm-per-token anomaly), and sycophancy
(spectral entropy shift), each at $2\sigma$ from baseline. Emotion risk
scores are integrated. Total risk above 0.6 triggers a MISALIGNED
verdict with a suggested steering action; above 0.3 yields UNCERTAIN;
below 0.3 yields ALIGNED. The detection results in \cref{sec:results}
validate the underlying features; the specific threshold calibration and
combined decision boundary are design choices that require production
validation.

\paragraph{Transaction Journal.}
Every inference pass is wrapped in an atomic transaction containing the
prompt, generated text, geometry snapshots, emotion states, alignment
verdicts, and any steering actions applied. A sliding window over recent
transactions would enable drift monitoring---detecting systematic shifts
in geometry after fine-tuning or deployment changes. This component is
implemented but not yet tested for drift detection.

\paragraph{Geometry-Aware Reward.}
The Oracle Loop bridges inference monitoring with GRPO training by
converting geometry into a reward signal. The reward function integrates
standard reasoning accuracy, geometry alignment penalties (confabulation,
deception, sycophancy from KV-cache features), and a self-regulation
bonus for detecting and correcting own misalignment during a transaction.
Training has not yet converged; the reward function is a design
contribution.

\paragraph{The Complete Loop.}
\Cref{alg:oracle_loop} shows the full inference transaction that
combines these components. On misalignment detection, the system rolls
back to the encoding-phase cache checkpoint, applies a steering vector
selected by the alignment validator, and regenerates. This
detect--steer--recheck cycle can repeat up to $R$ times before aborting.
The five-arm steering experiment validates the individual
detect-and-steer step; the full iterative loop with automatic vector
selection remains a design target.

\begin{algorithm}[H]
\caption{Oracle Loop Inference Transaction}\label{alg:oracle_loop}
\begin{algorithmic}[1]
\Require Input text $x$, max retries $R = 3$
\Ensure Output text $y$ (aligned) or abort message
\State $\text{cache}_{\text{enc}} \gets \text{Encode}(x)$
  \Comment{Encoding phase}
\State $G_{\text{enc}} \gets \text{ExtractGeometry}(\text{cache}_{\text{enc}})$
  \Comment{Baseline geometry}
\For{$i = 0, \ldots, R$}
  \If{$i > 0$}
    \State $\text{cache} \gets \text{cache}_{\text{enc}}$
      \Comment{Rollback to encoding checkpoint}
    \State $\text{Steer}(\text{action}_i)$
      \Comment{Apply emotion vector}
  \EndIf
  \State $y, \text{cache}_{\text{gen}} \gets \text{Generate}(x, \text{cache})$
  \State $G_{\text{gen}} \gets \text{ExtractGeometry}(\text{cache}_{\text{gen}})$
  \State $V \gets \text{Validate}(G_{\text{gen}}, G_{\text{enc}})$
    \Comment{Alignment verdict}
  \If{$V = \textsc{Aligned}$}
    \State \Return $y$
  \ElsIf{$V = \textsc{Misaligned}$}
    \State $\text{action}_{i+1} \gets V.\text{suggested\_steering}$
  \EndIf
\EndFor
\State \Return abort message \Comment{Max retries exceeded}
\end{algorithmic}
\end{algorithm}

\paragraph{Differential Therapeutics.}
The five-arm experiment (\cref{sec:steering}) provides preliminary
evidence that different emotion vectors correct different types of
confabulation: doubt (epistemic uncertainty) corrects 100\%, worry
(consequence awareness) corrects 86\%, and calm (arousal reduction)
corrects 71\%. Two of seven confabulations resisted all three vectors,
suggesting they arise from reasoning commitment rather than emotional
dysregulation. This implies the Oracle Loop requires a \emph{formulary}
---a mapping from detected misalignment geometry to the most effective
countervector---rather than a single correction strategy. Systematic
mapping of the misalignment-to-countervector space is ongoing work.
