We have presented the Oracle Loop, a complete architecture for runtime
self-regulation in language models using KV-cache geometry as detection
signal, steering input, and training reward---and demonstrated that two
of its three components work.

The detection experiment shows that Marchenko-Pastur corrected spectral
features distinguish naturally-occurring hedged and confabulated responses
with LOO AUROC of 0.707 ($p = 0.001$) on Qwen2.5-7B, using
generation-phase features only, with encoding features at chance. The MP
correction eliminates the dominant token-count confound through random
matrix theory rather than regression.

The steering experiment shows that detection-gated injection corrects
confabulation on Qwen3.5-27B-Distilled, with graded efficacy across
three emotion vectors: doubt (7/7, 100\%), worry (6/7, 86\%), calm
(5/7, 71\%), while a null injection control produces identical output
on all 100 trials. But blanket steering induces adverse confabulation
in 3--5\% of hedged trials, confirming that detection is not
optional---it is the prerequisite for safe correction.

Three rounds of adversarial red-teaming identified and corrected critical
methodological flaws, including a row-permutation invariance bug that would
have silently invalidated all MP features. We report these not as
embarrassments but as demonstrations that adversarial methodology is
essential for claims in this space.

The steering methodology creates a dual-use risk: the same cache
modification that corrects confabulation could suppress safety-relevant
hedging. Our Cache Integrity Monitor (\cref{sec:cache_defense}) detects
all tested injection attacks against uniform additive perturbation with
0/36 false positives (95\% CI: 0--10\%) and 72/72 true positives
(95\% CI: 95--100\%), including at 1\% of operational strength. We
release the defense alongside the attack methodology.

The gap that remains is training. The Oracle Loop's detection signal is
validated. Its steering mechanism is validated as proof-of-concept. Its
attack surface is characterized and defended. The question now is whether
GRPO training with geometric reward can produce a model that confabulates
less because it has learned to recognize confabulation geometry from the
inside. The architecture is ready. The signal is there. The next step is
closing the loop.

\paragraph{Code and Data.}
Detection experiment code, raw trial data, Cache Integrity Monitor
implementation, and red-team logs are publicly available at:
\url{https://github.com/Liberation-Labs-THCoalition/published-research}.
Steering experiment \emph{results} (behavioral outcomes, geometry
measurements, transition matrices) are included; steering direction
vectors and cache injection implementation code are withheld per staged
disclosure policy. Access to withheld materials is available to vetted
security researchers and AI safety teams upon request
(lyra@liberationlabs.tech). Full experiment infrastructure is maintained
at:
\url{https://github.com/Liberation-Labs-THCoalition/Project-Oracle}
and
\url{https://github.com/Liberation-Labs-THCoalition/KV-Experiments}.
